Investigating soundscapes perception through acoustic scenes simulation.

### 2017-07-31 lafaySimulatedScenePerception.tex
done with NM comments

  \begin{keyword}
  %% keywords here, in the form: keyword \sep keyword
  toto \sep toto
  %% PACS codes here, in the form: \PACS code \sep code

@@

First, a subject can be asked to describe a given sound following a given perceptive scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}. The amount of information that can be gathered strongly depends on the very nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}. That being said, those stimuli respectively are or can be recorded and analysed. It thus allows the experimenter to gather a coarse grain physical description of the stimuli.

@@

To the best of our knowledge, only Bruce~\al \cite{bruce2009development,bruce2014effects} considered the use of a simulator to question soundscape perception. They propose a tool that allows the user to modify a given soundscape by adding or removing specific sound sources, by changing the acoustic level of the sources as well as their spatial location. The authors shows that the addition or removal of the sources globally follows more social or semantic matters than their acoustical characteristics. A lack of diversity is nonetheless mentioned by the authors that limit the strenght of the outcomes given in this study.

@@

\emph{Simscene} is an online digital audio tool whose first version have been developed as part of the HOULE prokect\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}. It has been designed to run on the popular web browers \emph{Chrome} and \emph{Firefox}. It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{web-audio} standard that allows the manipulation of digita audio data within the browser\footnote{\emph{web-audio}: \url{www.w3.org/TR/webaudio}}. The interface for selecting the sound sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} \cite{d32011} visualization library.

@@

We choose not to include any musical content in the database of sounds, as the study of the pleasantness of a given style or genre of music is beyond the scope of the study.

@@

The simulation tool \emph{Simscene} follows this distinction and consider two distinct sound databases: one with the classes of events only and the other with classes of textures only. These two types also have specific scheduling procedures during the sumulation process.

@@

The event sequences are composed of salient events that can easily be recognized such as \emph{car start} or \emph{male speech}. They arise from a descriptive analysis based on the identification of the sound sources. On contrary, the amorphous sequences are sound environments where distinct events can hardly be found, such as \emph{traffic hubhub}. They build upon a holistic analysis based on global acoustic features.

@@

\begin{figure}[t]
        \myfloatalign
        \includegraphics[width=.8\linewidth]{gfx/ch_4/3.png}
       \caption{Hierachical organisation of the isolated sounds used ofr the simulation.}\label{fig:orgDb}
\end{figure}

@@

Once the typology and the set of sounds are available, an important design issue os the need of a suitable way to display the sound dataset to the user. Most browsing tools are based on keyword indexing. However, for sensory experiment that study the objectivation of the mental representation of the subject, this mean is problematic as the availability of a verbal description of the sound can influence the choice of the subject and potenitally induce biases in the analysis conclusions. Indeed, the subject may select \textit{a priori} sounds referenced as belonging to \emph{park} to build a \emph{calm} soundscape.

@@

Therefore, the selection interface considered in this study is text free and composed of spatially organized circles on a two dimensional plane.

@@

As such, the underlying model of the scene is a sum of sound sources, more precisely a "skeleton of events on a bed of textures" \cite{nelken2013ear}. Further details about the simulation interface can be found in \cite{rossignol2015simscene}.

@@

Experiment 1 aims at using a simulation paradigm to investigate the specific influences of the various sound sources constituting urban soundscapes on the perceived pleasantness. For that, the two first experiments are planned as follows (\cf~Figure~\ref{fig:xp1_2}):
\begin{itemize}
\item \emph{experiment 1.a (simulation)}: during this experiment, participants are asked to create urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simulator}. Each of them creates two environments: one ideal / pleasant and the other one, non ideal / unpleasant.
\item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes: respectively ideal (i) or non ideal (ni). Furthermore, this information is given by the creator of the scene. The second experimental step aims at investigating more deeply and more broadly our knowledge on pleasantness. For that, a second group of participants are asked to evaluate the pleasantness of each simulated scene, on a semantic scale. This experiment has two goals:
\begin{enumerate}
\item evaluate the respective influence on the pleasantness of the various sources composing the scenes (i or ni)
\item detect the presence of \emph{outliers} or ambiguous samples. For the remainder of the study, the defined hedonic properties (i and ni) are used as references. We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes; i.e that the lower i-scene is really above the upper ni-scene,  in terms of percieved pleasantness.
\end{enumerate}
\end{itemize}

@@

All the participants start by desiging the ideal environment; they read the second set of instruction at the end of the first experiment. Participants are completely free of their choice concerning sounds and synthesis parameters. They must nevertheless respect the two following constraints :
\begin{itemize}
\item the listening point of view is the one of a fixed listener;
\item the soundscape must be realistic, i.e. physically plausible.
\end{itemize}
For instance, the participant can insert ten dogs in the soundscape but she cannot insert one dog barking every 10 milliseconds.
These constraints are notified in the instructions; no control is done \emph{a priori} in the simulation user interface.

@@

A set of features is attached to each sound scene. The analysis is done on the basis of these features. A summary of these features (and the corresponding acronyms) is presented in Table ~\ref{tab:acronyme}. In order be consistent with the evaluation of experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version used as timuli for experiment 1.b, see Section \ref{sec:xp1b_plan}.
For each sound scene, three types of features are considered :
\begin{itemize}
\item \emph{perceptual}: the perceived pleasantness of the composed scene, assessed on a 7-point Likert scale. $\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the participants. $\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant. Given the low number of participants in Exp. 1.b, we choose not to normalize the pleasantness scores.
\item \emph{semantic}: a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene. Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise. The vector dimension ($n$) depends on the level of abstraction that is considered for analysis. For instance, for the abstraction level 1 that includes 44 classes of sounds, $n = 44$.
\item \emph{structural}: features related to \emph{sound level}. To figure out the sound level, we take the $L_{Aeq}$ feature as a conceptual reference. In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt. The level is obtained by computing the quadratic mean of the signal every second and averaging the results over the total duration of the scene. An A-filtering module processes the data before the quadratic means are computed. We note $L$, $L(E)$ and $L(T)$ the computed levels by respectively considering the whole set of samples, only the set of event samples and only the set of texture samples.
\end{itemize}

@@

\begin{itemize}
\item \emph{checking the pleasantness of the simulated sound scene}: to check that the emotional difference produced by the i- and ni- scenes is also present in terms of perceived pleasantness, we look at the existence of differences between $\mathcal{A}_{scene}$ and $\mathcal{A}_{subject}$ with respect to the two types of scenes. The significativity is evaluated by a Student test on two independent populations for $\mathcal{A}_{scene}$ and by a Student test on two paired populations for $\mathcal{A}_{subject}$;
\item \emph{analysis of the structural features}: in order to evaluate whether the affective difference between  the i- and ni- scenes has a significative influence  on the nature of the scenes \ie if significative differences exist between structural features and perceived pleasantness. The significativity is measured with a Student test on two populations;
\item \emph{study of the influence of the structural features on the perceived pleasantness}: in order to evaluate a possible impact of the structural features on the perceived pleasantness, linear correlations between these features and $\mathcal{A}_{scene}$ are considered. The Pearson correlation coefficient is used for that purpose, on the basis of the standard methodology used in multidimensional studies;
\item \emph{analysis of the semantic features}: in order to determine if the affective difference between  the i- and ni- scenes has an influence on the scenes composition in terms of sound sources, or if specific classes are more frequently used to simulate a given environment (i or ni), a V-test is considered. Testing is done at every level of abstraction, and separately for texture and event classes. For each class $j$, and each environment type $k$ ($k=\lbrace i,ni\rbrace$), the value $V_{jk}$ is computed as follows:

@@

where $\bigoplus$ is the \emph{exclusive-or}. The more the comosition of the two scenes are similar, the closer they are in this space. The use of Hamming distance allows us to take into account equally the presence and absence of classes. For measuring the intrinsic ability of the space to split the i- and ni- scene, we use a ranking metric named the precision at rank $k$ ($P@k$) . The $P@k$ computes the precision obtained after the $k$ closest items of a given seed item have been found. Formally, for each $s_i$ scene (considered as seed), we compute the ratio between the number of $s_j$ scenes taken among the $k$ nearest neighbors of $s_i$ and sharing the same label than $s_i$ versus $k$. the $P@k$ is then the average of this ratio for all the items considered as search seeds;

@@

It appears that the biggest influence on the global sound level comes from the events, the gap between $L$ and $L(E)$ being only 1 $dB$ for i- and ni- scenes. This observation is in agreement with the results obtained by Kuwano~\al~\cite{kuwano_memory_2003}. During their experiment, the authors ask to their participants to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source. The study shows that there is no significative difference between global and averaged instantaneous judgments. In our case, the result can be interpreted as if the participants had unconsciously integrated this perceptual reality when composing the scenes, by assigning the global sound level to well identified short sounds, \ie the events.

@@

The same conclusions can be drawn for $L(E)$, see~Figure~\ref{fig:soundlevele}. For $L(T)$, as shown on~Figure~\ref{fig:soundlevelf}, the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (i-scenes: $r=-0.33$, $p=0.05$, ni-scenes: $r=-0.00$, $p=0.99$). Again, we believe that the negative correlation coming from the whole is an artefact due to the level difference between the two types of scenes (i- scenes tend to be softer than ni - scenes). Thus, the sound events keep a relative ability to predict the pleasantness of the ni- scenes, but the level of the textures do not bring a lot of information whatever the type of environment is.

@@

Those first outcomes tend to show that 1) two modes of perception exists depending on the very nature of the environment (i- or ni-), and 2) each involve independent features. The fact that $L$ is not sufficient to characterize the pleasantness of the i- scenes can lead us to conclude that all the sound sources do not equally contribute to the perception of pleasantness. Thus, we think that only the level of some of them can influence this percept. In order to investigate further in that direction, we analyze in the next section the scenes on a semantic point of view, \ie we take an interest in the nature of the sources they are made of.

@@

We study the composition of the scenes by counting the number of participants who used a given class of sounds to simulate the environment. Results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures. For the sake of clarity, a transitional level of abstraction between level 0 and 1, named $0+$ is used to depict classes, see ~Figure~\ref{fig:taxonomie}.

@@

To a certain extent, our results are in contradiction with this assumption. In fact,  Figure~\ref{fig:soundsourcea} show that \emph{public transportation} classes (\emph{bus} and \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) have been used by the participants for $28\%$ of the i-scenes, and for $42\%$ of the ni-scenes. the results do not re-evaluate the fact that sounds of \emph{public transportation} are well accepted: $25\%$ of the subjects used \emph{bus} pour les i-scenes, a level that is comparable to the of the \emph{bike} class, and much higher than all the personal vehicles classes. Though, \emph{public transport} classes are also strongly present in the ni-scenes, for instance more than \emph{light vehicle} or \emph{trucks} classes. On the basis of our results, the public transportation class cannot be considered as typical of an ideal urban soundscape.

@@

Concerning $ET$, the $p@5$ is $76\%$ for the abstraction level $0$, and stay above $86\%$ from the abstraction level $1$. These results confirm that it is possible to make a clear distinction between the two types of environment on the sole basis of presence / absence of sound classes. We also notice that the higher the abstraction level is, the more important the ability to make this distinction is. In other words, the more precise in the scene description is, the more clear the discrimination between i- and ni- scenes is.

@@

We evaluate the correlations between $\mathcal{A}_{scene}$ and the structural features. In this section, the structural features are computed by taking into account only the previously identified sound markers. We define $Xm$, the $X$ feature computed by taking into account only the sound markers. On the opposite, we define $X_b$, the $X$ feature computed by taking into account all the sound classes, except the sound markers. When the feature characterizes an i- scene (idem for a ni- scene),  only the markers identified for the i- scenes (or the ni- scenes) are considered; we named them, i- markers (or ni- markers). Results are shown in Table~\ref{tab:corrMarkers}.

@@

Considering the non markers classes, we can observe on the i- scenes results, a weak negative correlation for $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$), see Figures~\ref{fig:soundlevelNoisea}  and ~\ref{fig:soundlevelNoised}. It is the first time that an objective feature allows us to define the pleasantness of the pleasant environments. This leads us to conclude that the level of non-typical sound classes of a pleasant environment has a negative influence on the pleasantness.

@@

We consider a last group of features, namely $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$.  These features describe the difference between the markers level and those of the other sound classes, see ~Figures~\ref{fig:soundlevelMarkerDiffa} and~\ref{fig:soundlevelMarkerDiffd}). They express the saliency of the markers with respect to the sound mixture.

@@

\begin{itemize}
\item \emph{differentiating the i- and ni-scenes}: the semantic features, and the global structural features ($L$, $L(E)$ et $L(T)$), allow us to differentiate between i- and ni- scenes. The semantic description seems to be more powerful;
\item \emph{events or textures}: whatever the feature type, be it semantic or structural, events are mainly useful to differentiate the two types of events; textures bring a limited amount of information;
\item \emph{pleasantness prediction}: considering the correlation between structural features and pleasantness, it seems that the way subjects perceived the quality of a given environment depends on its nature (i or ni). From the data gathred in those experiments, it is not conceivable to consider the same set of features to predict the pleasantness of i- and ni- scenes in the same time:

@@

The fact that the pleasantness of the i- scenes is not correlated to global physical features, contrary to pleasantness of ni- scenes has been studied recently \cite{gozalo2015relationship}.

We thus observe two perceptual modes of operation that involve different types of features, and that rely on the hedonic nature  of the stimuli. A phenomenon similar is observed for the perception of textures, see Section~\ref{sec:simscene_sampleDataSet}. It seems that the brain adapts its way to encode the information (statistic summary for textures, fine description for events) following a previous decision making with regards to the nature of the stimuli, \ie is it an event or a texture ?. Similarly, in the light of this study, it appears that the features considered for the pleasantness judgment also depends on a preliminary judgment of the global hedonic nature of the environment (ideal or non ideal).

@@

In order to verify that the pleasantness of the ideal and non ideal scenes is affected by the sound markers, the audio waveform of the simulated scenes are regenerated with or without using the classes identified as markers. For the i-scenes, some of the i-markers are removed. For the ni-scenes, some of the ni-markers are removed. To evaluate the impact on the perception of pleasantness caused by those modifications, a perceptive test is then conducted with a protocol close to the one considered in experiment 1.b.

@@

Finally, these works allow us to conjecture as to the nature of the mental representations of the concepts "pleasant urban sound environment" (EA) and "unpleasant urban sound environment" (ED).

@@

First, the fact that the semantic information (in terms of which sound source is present) and compositional information are different for ideal and non-ideal scenes leads us to believe that these two types of information characterize the EA and ED concepts.

@@

Secondly, the fact that the removal of sound markers modifies the perceived pleasantness leads us to believe that the abstract concept related to pleasantness depends on the activation of a network of concrete concepts strongly linked to the sources which are in the case of this study: \emph{bird}, \emph{church bell} and \emph{bicycle bell}).

@@

Third, the fact that the compositional descriptors correlated with the perceived pleasantness change according to the type of scenes leads us to believe that the compositional characteristics that allow us to distinguish between two instances of the same abstract concept linked to the pleasantness are not the same for EA and ED.

@@

Future work should consider several other simulation experiments by varying the emotional qualities (quiet, comfortable, troublesome, etc.), but also by specifying specific urban places (park, square, street, etc.), in order to provide to the scientific community an entire corpora of cognitively informed soundscapes.

@@

\begin{figure}[h]
        \myfloatalign
        \subfloat[]
        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/event_1_en}\label{fig:taxonomieEventa}}
        \subfloat[]
        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/event_2_en}\label{fig:taxonomieEventb}}\par
        \subfloat[]
        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/texture_1_en}\label{fig:taxonomieTexturea}}
        \subfloat[]
        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/texture_2_en}\label{fig:taxonomieTextureb}}
       \caption{Taxonomy of sound classes used for the simulation urban soundscapes: events (a,b) and textures (c,d).}
       \label{fig:taxonomie}
\end{figure}

>>>>

  \begin{keyword}
  %% keywords here, in the form: keyword \sep keyword
 cognitive psychology \sep soundscape assessment
  %% PACS codes here, in the form: \PACS code \sep code

@@

First, a subject can be asked to describe a given sound following a given perceptual scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}. The amount of information that can be gathered strongly depends on the very nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}. That being said, those stimuli respectively are or can be recorded and analysed. It thus allows the experimenter to gather a coarse grain physical description of the stimuli.

@@

To the best of our knowledge, only Bruce~\cite{bruce2009development,bruce2014effects} considered the use of a simulator to question soundscape perception. They propose a tool that allows the user to modify a given soundscape by adding or removing specific sound sources, by changing the acoustic level of the sources as well as their spatial location. The authors shows that the addition or removal of the sources globally follows more social or semantic matters than their acoustical characteristics. A lack of diversity is nonetheless mentioned by the authors that limit the strenght of the outcomes given in this study.

@@

\emph{Simscene} is an online digital audio tool whose first version have been developed as part of the HOULE project\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}. It has been designed to run on the popular web browsers \emph{Chrome} and \emph{Firefox}. It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{web-audio} standard that allows the manipulation of digita audio data within the browser\footnote{\emph{web-audio}: \url{www.w3.org/TR/webaudio}}. The interface for selecting the sound sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} \cite{d32011} visualization library.

@@

We choose not to include any musical content in the sound database, as the study of the pleasantness of a given style or genre of music is beyond the scope of the study.

@@

The simulation tool \emph{Simscene} follows this distinction and consider two distinct sound databases: one with the classes of events only and the other with classes of textures only. These two types also have specific scheduling procedures during the simulation process.

@@

The event sequences are composed of salient events that can easily be recognized such as \emph{car start} or \emph{male speech}. They arise from a descriptive analysis based on the identification of the sound sources. On contrary, the amorphous sequences are sound environments where distinct events can hardly be found, such as \emph{traffic hubhub}. They result from a holistic analysis based on global acoustic features.

@@

\begin{figure}[t]
        \myfloatalign
        \includegraphics[width=.8\linewidth]{gfx/ch_4/3.png}
       \caption{Hierachical organisation of the isolated sounds used of the simulation.}\label{fig:orgDb}
\end{figure}

@@

Once the typology and the set of sounds are available, an important design issue is the need of a suitable way to display the sound dataset to the user. Most browsing tools are based on keyword indexing. However, for sensory experiment that study the objectivation of the mental representation of the subject, this mean is problematic as the availability of a verbal description of the sound can influence the choice of the subject and potenitally induce biases in the analysis conclusions. Indeed, the subject may select \textit{a priori} sounds referenced as belonging to \emph{park} to build a \emph{calm} soundscape.

@@

Therefore, the selection interface considered in this study is text free and composed of spatially organized circles on a plane.

@@

As such, the underlying model of the scene is a sum of sound sources. Further details about the simulation interface can be found in \cite{rossignol2015simscene}.

@@

Experiment 1 aims at using a simulation paradigm to investigate the specific influences of the various sound sources constituting urban soundscapes on the perceived pleasantness. For that, the two first experiments are planned as follows (\cf~Figure~\ref{fig:xp1_2}):
\begin{itemize}
\item \emph{experiment 1.a (simulation)}: during this experiment, participants are asked to create urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simulator}. Each of them creates two environments: one ideal / pleasant and the other one, non ideal / unpleasant.
\item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes: respectively ideal (i) or non ideal (ni). Furthermore, this information is given by the creator of the scene. The second experimental step aims at investigating more deeply and more broadly our knowledge on pleasantness. For that, a second group of participants are asked to evaluate the pleasantness of each simulated scene, on a semantic scale. This experiment has two goals:
\begin{enumerate}
\item evaluate the respective influence on the pleasantness of the various sources composing the scenes (i or ni)
\item detect the presence of \emph{outliers} or ambiguous samples. For the remainder of the study, the defined hedonic properties (i and ni) are used as references. We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes; i.e that the lower i-scene is really above the upper ni-scene,  in terms of perceived pleasantness.
\end{enumerate}
\end{itemize}

@@

All the participants start by designing the ideal environment; they read the second set of instruction at the end of the first experiment. Participants are completely free of their choice concerning sounds and synthesis parameters. They must nevertheless respect the two following constraints :
\begin{itemize}
\item the listening point of view is the one of a fixed listener;
\item the soundscape must be realistic, i.e. physically plausible.
\end{itemize}
For instance, the participant can insert ten dogs in the soundscape but she cannot insert one dog barking every 10 milliseconds.
These constraints are notified in the instructions; no control is done \emph{a priori} in the simulation user interface.

@@

A set of features is attached to each sound scene. The analysis is done on the basis of these features. A summary of these features (and the corresponding acronyms) is presented in Table ~\ref{tab:acronyme}. In order be consistent with the evaluation of experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b, see Section \ref{sec:xp1b_plan}.
For each sound scene, three types of features are considered :
\begin{itemize}
\item \emph{perceptual}: the perceived pleasantness of the composed scene, assessed on a 7-point Likert scale. $\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the participants. $\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant. Given the low number of participants in Exp. 1.b, we choose not to normalize the pleasantness scores.
\item \emph{semantic}: a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene. Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise. The vector dimension ($n$) depends on the level of abstraction that is considered for analysis. For instance, for the abstraction level 1 that includes 44 classes of sounds, $n = 44$.
\item \emph{structural}: features related to \emph{sound level}. To figure out the sound level, we take the $L_{Aeq}$ feature as a conceptual reference. In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt. The level is obtained by computing the quadratic mean of the signal every second and averaging the results over the total duration of the scene. An A-filtering module processes the data before the quadratic means are computed. We note $L$, $L(E)$ and $L(T)$ the computed levels by respectively considering the whole set of samples, only the set of event samples and only the set of texture samples.
\end{itemize}

@@

\begin{itemize}
\item \emph{checking the pleasantness of the simulated sound scene}: to check that the emotional difference produced by the i- and ni- scenes is also present in terms of perceived pleasantness, we look at the existence of differences between $\mathcal{A}_{scene}$ and $\mathcal{A}_{subject}$ with respect to the two types of scenes. The significativity is evaluated by a Student test on two independent populations for $\mathcal{A}_{scene}$ and by a Student test on two paired populations for $\mathcal{A}_{subject}$;
\item \emph{analysis of the structural features}: in order to evaluate whether the affective difference between  the i- and ni- scenes has a significative influence  on the nature of the scenes \ie~ if significative differences exist between structural features and perceived pleasantness. The significativity is measured with a Student test on two populations;
\item \emph{study of the influence of the structural features on the perceived pleasantness}: in order to evaluate a possible impact of the structural features on the perceived pleasantness, linear correlations between these features and $\mathcal{A}_{scene}$ are considered. The Pearson correlation coefficient is used for that purpose, on the basis of the standard methodology used in multidimensional studies;
\item \emph{analysis of the semantic features}: in order to determine if the affective difference between  the i- and ni- scenes has an influence on the scenes composition in terms of sound sources, or if specific classes are more frequently used to simulate a given environment (i or ni), a V-test is considered. Testing is done at every level of abstraction, and separately for texture and event classes. For each class $j$, and each environment type $k$ ($k=\lbrace i,ni\rbrace$), the value $V_{jk}$ is computed as follows:

@@

where $\bigoplus$ is the \emph{exclusive-or}. The more the composition of the two scenes are similar, the closer they are in this space. The use of Hamming distance allows us to take into account equally the presence and absence of classes. For measuring the intrinsic ability of the space to split the i- and ni- scene, we use a ranking metric named the precision at rank $k$ ($P@k$) . The $P@k$ computes the precision obtained after the $k$ closest items of a given seed item have been found. Formally, for each $s_i$ scene (considered as seed), we compute the ratio between the number of $s_j$ scenes taken among the $k$ nearest neighbors of $s_i$ and sharing the same label than $s_i$ versus $k$. the $P@k$ is then the average of this ratio for all the items considered as search seeds;

@@

It appears that the biggest influence on the global sound level comes from the events, the gap between $L$ and $L(E)$ being only 1 $dB$ for i- and ni- scenes. This observation is in agreement with the results obtained by Kuwano~\al~\cite{kuwano_memory_2003}. During their experiment, the authors ask their participants to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source. The study shows that there is no significative difference between global and averaged instantaneous judgments. In our case, the result can be interpreted as if the participants had unconsciously integrated this perceptual reality when composing the scenes, by assigning the global sound level to well identified short sounds, \ie the events.

@@

The same conclusions can be drawn for $L(E)$, see~Figure~\ref{fig:soundlevele}. For $L(T)$, as shown on~Figure~\ref{fig:soundlevelf}, the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (i-scenes: $r=-0.33$, $p=0.05$, ni-scenes: $r=-0.00$, $p=0.99$). Again, we believe that the negative correlation coming from the whole is an artefact due to the level difference between the two types of scenes (i- scenes tend to be softer than ni- scenes). Thus, the sound events keep a relative ability to predict the pleasantness of the ni- scenes, but the level of the textures do not bring a lot of information whatever the type of environment is.

@@

Those first outcomes tend to show that 1) two modes of perception exist depending on the very nature of the environment (i- or ni-), and 2) each involve independent features. The fact that $L$ is not sufficient to characterize the pleasantness of the i- scenes can lead us to conclude that all the sound sources do not equally contribute to the perception of pleasantness. Thus, we think that only the level of some of them can influence this percept. In order to investigate further in that direction, we analyze in the next section the scenes on a semantic point of view, \ie we take an interest in the nature of the sources they are made of.

@@

We study the composition of the scenes by counting the number of participants who used a given class of sounds to simulate the environment. Results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures. For the sake of clarity, a transitional level of abstraction between level 0 and 1, named $0+$ is used to depict classes, see ~Figures~\ref{fig:taxonomieEa}, \ref{fig:taxonomieEb} and \ref{fig:taxonomieT}.

@@

To a certain extent, our results are in contradiction with this assumption. In fact,  Figure~\ref{fig:soundsourcea} show that \emph{public transportation} classes (\emph{bus} and \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) have been used by the participants for $28\%$ of the i-scenes, and for $42\%$ of the ni-scenes. the results do not re-evaluate the fact that sounds of \emph{public transportation} are well accepted: $25\%$ of the subjects used \emph{bus} for the i-scenes, a level that is comparable to the \emph{bike} class, and much higher than all the personal vehicles classes. Though, \emph{public transport} classes are also strongly present in the ni-scenes, for instance more than \emph{light vehicle} or \emph{trucks} classes. On the basis of our results, the public transportation class cannot be considered as typical of an ideal urban soundscape.

@@

Concerning $ET$, the $p@5$ is $76\%$ for the abstraction level $0$, and stay above $86\%$ for the abstraction level $1$. These results confirm that it is possible to make a clear distinction between the two types of environment on the sole basis of presence / absence of sound classes. We also notice that the higher the abstraction level is, the more important the ability to make this distinction is. In other words, the more precise the scene description is, the more clear the discrimination between i- and ni- scenes is.

@@

We evaluate the correlations between $\mathcal{A}_{scene}$ and the structural features. In this section, the structural features are computed by taking into account only the previously identified sound markers. We define $X_m$, the $X$ feature computed by taking into account only the sound markers. On the opposite, we define $X_b$, the $X$ feature computed by taking into account all the sound classes, except the sound markers. When the feature characterizes an i- scene (idem for a ni- scene),  only the markers identified for the i- scenes (or the ni- scenes) are considered; we named them, i- markers (or ni- markers). Results are shown in Table~\ref{tab:corrMarkers}.

@@

Considering the non markers classes, we can observe on the i- scenes results, a weak negative correlation for $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$), see Figures~\ref{fig:soundlevelNoisea}  and ~\ref{fig:soundlevelNoised}. It is the first time that an objective feature allows us to define the pleasantness of i- scenes. This leads us to conclude that the level of non-typical sound classes of a pleasant environment has a negative influence on the pleasantness.

@@

We consider a last group of features, namely $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$.  These features describe the difference between the markers level and those of the other sound classes, see ~Figures~\ref{fig:soundlevelMarkerDiffa} and~\ref{fig:soundlevelMarkerDiffd}. They express the saliency of the markers with respect to the sound mixture.

@@

\begin{itemize}
\item \emph{differentiating the i- and ni-scenes}: the semantic features, and the global structural features ($L$, $L(E)$ et $L(T)$), allow us to differentiate between i- and ni- scenes. The semantic description seems to be more powerful;
\item \emph{events or textures}: whatever the feature type, be it semantic or structural, events are mainly useful to differentiate the two types of scenes; textures bring a limited amount of information;
\item \emph{pleasantness prediction}: considering the correlation between structural features and pleasantness, it seems that the way subjects perceived the quality of a given environment depends on its nature (i or ni). From the data gathered in those experiments, it is not conceivable to consider the same set of features to predict the pleasantness of i- and ni- scenes in the same time:

@@

The fact that the pleasantness of the i- scenes is not correlated to global physical features, contrary to pleasantness of ni- scenes, has been studied recently \cite{gozalo2015relationship}.

We observe two perceptual modes of operation that involve different types of features, and that rely on the hedonic nature  of the stimuli. A similar phenomenon is observed for the perception of textures, see Section~\ref{sec:simscene_sampleDataSet}. It seems that the brain adapts its way to encode the information (statistic summary for textures, fine description for events) following a previous decision making with regards to the nature of the stimuli, \ie~is it an event or a texture ? Similarly, in the light of this study, it appears that the features considered for the pleasantness judgment also depends on a preliminary judgment of the global hedonic nature of the environment (ideal or non ideal).

@@

In order to verify that the pleasantness of the ideal and non ideal scenes is affected by the sound markers, the audio waveform of the simulated scenes are regenerated with or without the classes identified as markers. For the i-scenes, some of the i-markers are removed. For the ni-scenes, some of the ni-markers are removed. To evaluate the impact on the perception of pleasantness caused by those modifications, a perceptual test is then conducted with a protocol close to the one considered in experiment 1.b.

@@

Finally, these works allow us to conjecture as to the nature of the mental representations of the concepts "pleasant (urban sound) environment" (PE) and "unpleasant (urban sound) environment" (UE).

@@

First, the fact that the semantic information (in terms of which sound source is present) and compositional information are different for ideal and non-ideal scenes leads us to believe that these two types of information characterize the PE and UE concepts.

@@

Secondly, the fact that the removal of sound markers modifies the perceived pleasantness leads us to believe that the abstract concept related to pleasantness depends on the activation of a network of concrete concepts strongly linked to the sources which are in the case of this study: \emph{bird}, \emph{church bell} and \emph{bicycle bell}.

@@

Third, the fact that the compositional descriptors correlated with the perceived pleasantness change according to the type of scenes leads us to believe that the compositional characteristics that allow us to distinguish between two instances of the same abstract concept linked to the pleasantness are not the same for PE and UE.

@@

Future work should consider several other simulation experiments by varying the emotional qualities (quiet, comfortable, troublesome, etc.), but also by specifying specific urban locations (park, square, street, etc.), in order to provide to the scientific community an entire corpora of cognitively informed soundscapes.

@@

\begin{figure}[h]
    \centering
        {\includegraphics[trim={ 0 0 0cm 1cm},clip,width=.7\columnwidth]{gfx/ch_5/event_1_en}\label{fig:taxonomieEventa}} \par
         \caption{Taxonomy of sound classes of mechanical events used for the simulation urban soundscapes  with level of abstraction from left to right.}
       \label{fig:taxonomieEa}
\end{figure}

\begin{figure}[h]
    \centering
        {\includegraphics[trim={ 0 0 0cm 1cm},clip,width=.7\columnwidth]{gfx/ch_5/event_2_en}\label{fig:taxonomieEventb}}
       \caption{Taxonomy of sound classes of non-mechanical events used for the simulation urban soundscapes with level of abstraction from left to right.}
       \label{fig:taxonomieEb}
\end{figure}

\begin{figure}[h]
  \centering
        {\includegraphics[trim={ 0 0 7cm 0},clip,width=.7\columnwidth]{gfx/ch_5/texture_1_en}\label{fig:taxonomieTexturea}} \par

        {\includegraphics[trim={ 0 0 7cm 1cm},clip,width=.7\columnwidth]{gfx/ch_5/texture_2_en}\label{fig:taxonomieTextureb}}
       \caption{Taxonomy of sound classes of textures used for the simulation urban soundscapes with level of abstraction from left to right.}
       \label{fig:taxonomieT}
\end{figure}
